---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Mixture of Markov Chains with Support of Higher Orders and Multiple Sequences

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/markovmix)](https://CRAN.R-project.org/package=markovmix)
[![R-CMD-check](https://github.com/zhuxr11/markovmix/workflows/R-CMD-check/badge.svg)](https://github.com/zhuxr11/markovmix/actions)
[![Codecov test coverage](https://codecov.io/gh/zhuxr11/markovmix/branch/master/graph/badge.svg)](https://app.codecov.io/gh/zhuxr11/markovmix?branch=master)
<!-- badges: end -->

**Package**: [*markovmix*](https://github.com/zhuxr11/markovmix) `r pkgload::pkg_version()`<br />
**Author**: Xiurui Zhu<br />
**Modified**: `r file.info("README.Rmd")$mtime`<br />
**Compiled**: `r Sys.time()`

The goal of `markovmix` is to fit mixture of Markov chains of higher orders from multiple sequences.
It is also compatible with ordinary 1-component, 1-order or single-sequence Markov chains.
Various utility functions are defined to derive transition patterns, transition probabilities per component
and component priors. In addition, `print()` and `predict()` methods are also provided as a convention of model fitting.

## Installation

You can install the released version of `markovmix` from [CRAN](https://cran.r-project.org/) with:

``` r
install.packages("markovmix")
```

Alternatively, you can install the developmental version of `markovmix` from [github](https://github.com/) with:

``` r
remotes::install_github("zhuxr11/markovmix")
```

## Examples of fitting (mixture of) Markov chains

Before start, we generate some sequences with a finite set of states, just as those to fit a Markov chain.

```{r library, eval=FALSE}
library(markovmix)
```

```{r, include=FALSE}
pkgload::load_all()
```

```{r data-prep}
library(purrr)

# Define Markov states
mk_states <- LETTERS[seq_len(4L)]
# Define number of sequences
mk_size <- 100L
# Define sequence length range
mk_len_range <- seq_len(10L)
# Define a helper function to generate sequence list
gen_seq_list <- function(size, len_range, states, seed = NULL) {
  if (is.null(seed) == FALSE) {
    set.seed(seed)
  }
  purrr::map(seq_len(size), ~ sample(states, sample(len_range, 1L), replace = TRUE))
}
mk_seq_list <- gen_seq_list(size = mk_size, len_range = mk_len_range, states = mk_states, seed = 1111L)
head(mk_seq_list)
```

First, we fit a 1-order Markov chain, the most commonly used type.

```{r}
mk_fit <- fit_markov_mix(seq_list = mk_seq_list, states = mk_states)
print(mk_fit)
```

Then, we fit a 2-order Markov chain.

```{r}
mk_fit2 <- fit_markov_mix(seq_list = mk_seq_list, order. = 2L, states = mk_states)
print(mk_fit2)
```

Finally, we fit a 3-component mixture of 2-order Markov chains, with randomly generated soft clustering probabilities for each sequence.

```{r}
# Define number of components
mk_n_comp <- 3L
set.seed(2222L)
mk_clusters <- matrix(runif(length(mk_seq_list) * mk_n_comp), nrow = length(mk_seq_list), ncol = mk_n_comp)
mk_mix_fit2 <- fit_markov_mix(seq_list = mk_seq_list, order. = 2L, states = mk_states, clusters = mk_clusters)
print(mk_mix_fit2)
```

## Examples of predicting (mixture of) Markov chains

Before start, we generate some sequences for prediction.

```{r data-prep-pred}
mk_pred_seq_list <- gen_seq_list(size = 50L, len_range = mk_len_range, states = mk_states, seed = 3333L)
```

We predict the probabilities of each sequence in the generated mixture of Markov chains.

```{r}
mk_pred_res <- predict(mk_mix_fit2, newdata = mk_pred_seq_list)
print(mk_pred_res)
```

Please note that `NA` values are assigned to the probabilities of sequences if no valid sub-sequences (such as those of length `r mk_mix_fit2[["order"]] + 1` or longer).

```{r}
print(purrr::map_int(mk_pred_seq_list[is.na(mk_pred_res) == TRUE], length))
```

To predict the sequences in each component of the mixture of Markov chains, we can use `aggregate. = FALSE`.

```{r}
mk_pred_res_comp <- predict(mk_mix_fit2, newdata = mk_pred_seq_list, aggregate. = FALSE)
print(head(mk_pred_res_comp, n = 10L))
```

## Time of computation

The core functions of `markovmix` are written with `Rcpp` to efficiently process sequences and probabilities. Here, time of computation (ToC) is profiled with the number of sequences, where dots correspond to ToC medians and error bars to ToC quantiles.

```{r toc, cache=TRUE}
# Define sequence lengths for ToC profiling
mk_toc_seq_len <- as.integer(outer(c(1L, 2L, 5L), 10^(seq_len(5L) - 1L)))
set.seed(4444L)
mk_toc_seq_list <- purrr::map(
  mk_toc_seq_len,
  ~ gen_seq_list(size = .x, len_range = mk_len_range, states = mk_states)
)
mk_toc_clusters <- purrr::map(
  mk_toc_seq_len,
  ~ matrix(runif(.x * mk_n_comp), nrow = .x, ncol = mk_n_comp)
)
mk_toc_exprs <- purrr::map(
  purrr::set_names(seq_along(mk_toc_seq_len), mk_toc_seq_len),
  ~ rlang::expr(fit_markov_mix(
    seq_list = mk_toc_seq_list[[!!.x]],
    order. = 2L,
    states = mk_states,
    clusters = mk_toc_clusters[[!!.x]],
    verbose = FALSE
  ))
)
# Profile time of computation
mk_toc <- bench::mark(
  exprs = mk_toc_exprs,
  min_time = Inf,
  iterations = 5L,
  filter_gc = FALSE,
  check = FALSE
)
```

```{r toc-plots}
mk_toc_plot_data <- mk_toc %>%
  tibble::as_tibble() %>%
  dplyr::select(expression, time) %>%
  dplyr::rename(seq_len = expression, toc = time) %>%
  dplyr::mutate_at("seq_len", ~ as.integer(names(.x))) %>%
  dplyr::mutate_at("toc", ~ purrr::map(.x, unclass)) %>%
  tidyr::unnest("toc")
ggplot2::ggplot(mk_toc_plot_data %>%
                  dplyr::group_by(seq_len) %>%
                  dplyr::summarize(toc_median = median(toc, na.rm = TRUE),
                                   toc_min = quantile(toc, 0.25, na.rm = TRUE),
                                   toc_max = quantile(toc, 0.75, na.rm = TRUE),
                                   .groups = "drop"),
                ggplot2::aes(x = seq_len, y = toc_median)) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = toc_min, ymax = toc_max)) +
  ggplot2::scale_x_log10(breaks = mk_toc_seq_len) +
  ggplot2::scale_y_log10() +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, vjust = 1)) +
  ggplot2::labs(x = "Number of sequences", y = "Time of computation (s)")
```

